---
title: "KL Mammal Occupancy Paper - Revisions 2"
author: "Joseph Redinger"
date: "9/8/2023"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

### **Script Summary:** Fits python, year, and environmental factors to occupancy models to mammal camera trap data. Creates figures to help visualize changes in detection and extinction across 2016-2020.

### **Paper:** Redinger J, Lord I, Dixon J, and Cove MV (2023) Mammal declines correspond with the increasing prevalence of Burmese pythons at their southern invasion front in the Florida Keys. Biological Invasions (In Review)

# Occupancy Models

### Load libraries
```{r, warning = FALSE}
library(unmarked)  # create occupancy models
library(ggplot2)   # create individual plots
library(ggpubr)    # combine plots for publication
library(dplyr)     # clean data frame
```

### Load Data
```{r, results="hide"}
# load environmental variables - only dist.to.development used in models
env.covs = read.csv("CM_occ_dat_covs.csv", header = T)

# load python variables
python.covs = read.csv("CLNWR_eMammal_python_variables.csv", header = T) 

# load mammal occupancy data
raccoon = read.csv("Northern_Raccoon_detection_history.csv", header = T) 
opossum = read.csv("Virginia_Opossum_detection_history.csv", header = T)  
squirrel = read.csv("Eastern_Gray_Squirrel_detection_history.csv", header = T)
cotton.mouse = read.csv("Cotton_Mouse_detection_history.csv", header = T)
woodrat = read.csv("Eastern_Woodrat_detection_history.csv", header = T)
black.rat = read.csv("House_rat_detection_history.csv", header = T)

```

### Create function that prepares data for unmarked usage
```{r, warning = FALSE}
format.occ.dat = function(x) {
  mam.2016 = x[which(x$year == 2016), 3:10]
  mam.2018 = x[which(x$year == 2018), 3:10]
  mam.2020 = x[which(x$year == 2020), 3:10]
  mam.dat = merge(env.covs, python.covs, by = "Camera.ID")
  mam.dat = merge(mam.dat, mam.2016, by = "Camera.ID")
  mam.dat = merge(mam.dat, mam.2018, by = "Camera.ID")
  mam.dat = merge(mam.dat, mam.2020, by = "Camera.ID")
  return(mam.dat)
}

```


### Function to run models
```{r, warning = FALSE}
run.occ.models = function(data) {
  mam.dat = format.occ.dat(data)
  survey.years = matrix(c('2016','2018','2020'), 428, 3, byrow = T)

  mam.time.umf = unmarkedMultFrame(y = mam.dat[27:47],
                                      siteCovs = mam.dat[1:26],
                                      yearlySiteCovs = list(year = survey.years),
                                      numPrimary = 3)

  occ.0 = colext(~1, ~1, ~1, ~1, mam.time.umf)
  occ.1 = colext(~1, ~1, ~1, ~year-1, mam.time.umf)
  occ.2 = colext(~1, ~1, ~1, ~scale(dist.to.development), mam.time.umf)
  occ.3 = colext(~1, ~1, ~year-1, ~1, mam.time.umf)
  occ.4 = colext(~1, ~1, ~year-1, ~scale(dist.to.development), mam.time.umf)
  occ.5 = colext(~1, ~1, ~year-1, ~year-1, mam.time.umf)
  occ.6 = colext(~1, ~1, ~year-1, ~year-1 + scale(dist.to.development), mam.time.umf)
  
  fl = fitList(occ.0, occ.1, occ.2, occ.3, occ.4, occ.5, occ.6)
  ms_occ = modSel(fl)
  ms_occ

  mam.summary = list(ms_occ, get(ms_occ@Full$model[1]), ms_occ@Full)
  return(mam.summary)
}

```

#### Raccoon Models
```{r, warning = FALSE}
ra.mods = run.occ.models(raccoon)
ra.mods[1:2]


```

#### Opossum Models
```{r, warning = FALSE}
op.mods = run.occ.models(opossum)
op.mods[1:2]

```

#### Squirrel Models
```{r, warning = FALSE}
sq.mods = run.occ.models(squirrel)
sq.mods[1:2]

```

#### Cotton Mouse Models
```{r, warning = FALSE}
cm.mods = run.occ.models(cotton.mouse)
cm.mods[1:2]

```

#### Woodrat Models
```{r, warning = FALSE}
wr.mods = run.occ.models(woodrat)
wr.mods[1:2]

```

#### Blackrat models
```{r, warning = FALSE}
br.mods = run.occ.models(black.rat)
br.mods[1:2]

```
 
### Combine occupancy model results into table
```{r, warning = FALSE, echo=FALSE}
ra.tab = as.data.frame(ra.mods[3]) %>%
  filter(row_number() == 1 | cumltvWt < 0.9)
op.tab = as.data.frame(op.mods[3]) %>%
  filter(row_number() == 1 | cumltvWt < 0.9)
sq.tab = as.data.frame(sq.mods[3]) %>%
  filter(row_number() == 1 | cumltvWt < 0.9)
cm.tab = as.data.frame(cm.mods[3]) %>%
  filter(row_number() == 1 | cumltvWt < 0.9)
wr.tab = as.data.frame(wr.mods[3]) %>%
  filter(row_number() == 1 | cumltvWt < 0.9)
br.tab = as.data.frame(br.mods[3]) %>%
  filter(row_number() == 1 | cumltvWt < 0.9)

table = rbind(ra.tab, op.tab, sq.tab, cm.tab, wr.tab, br.tab)

mam.species = c(rep("raccoon", nrow(ra.tab)), rep("opossum", nrow(op.tab)), 
             rep("squirrel", nrow(sq.tab)), rep("cotton.mouse", nrow(cm.tab)),  
             rep("woodrat", nrow(wr.tab)), rep("black.rat", nrow(br.tab)))
table$species = mam.species
table
 

# selecting columns
col_order = c("species", "model", "formula", "nPars", "delta", "cumltvWt", "ext.Int.", 
              "SEext.Int.", "ext.year2016.", "SEext.year2016.", "ext.year2018.",
              "SEext.year2018.", "p.Int.", "SEp.Int.", "p.scale.dist.to.development..",
              "SEp.scale.dist.to.development..", "p.year2016.", "SEp.year2016.",
              "p.year2018.", "SEp.year2018.", "p.year2020.", "SEp.year2020." )
table2= table[, col_order]
table2

# round columns to decimal places
round = as.data.frame(apply(table2[5:ncol(table2)], 2, function(x) ifelse(is.na(x), NA, round(x, digits = 2))))
table3 = cbind(table2[1:4], round)

# combine columns
table3$extIntercept = paste(table3$ext.Int., "±", table3$SEext.Int.)
table3$ext2016 = paste(table3$ext.year2016., "±", table3$SEext.year2016.)
table3$ext2018 = paste(table3$ext.year2018., "±", table3$SEext.year2018.)
table3$pIntercept = paste(table3$p.Int., "±", table3$SEp.Int.)
table3$dev = paste(table3$p.scale.dist.to.development.., "±", table3$SEp.scale.dist.to.development..)
table3$p2016 = paste(table3$p.year2016., "±", table3$SEp.year2016.)
table3$p2018 = paste(table3$p.year2018., "±", table3$SEp.year2018.)
table3$p2020 = paste(table3$p.year2020., "±", table3$SEp.year2020.)
table3$pIntercept = paste(table3$ext.Int., "±", table3$SEext.Int.)

# selecting clean columns
new_col_order = c("species", "formula", "nPars", "cumltvWt", "delta", "extIntercept", 
                  "ext2016", "ext2018", "pIntercept", "dev", "p2016", "p2018","p2020" )
table.clean= table3[, new_col_order]
table.clean

table.clean[table.clean == "NA ± NA"] = "-"
# rename column name
table.clean = table.clean %>%
  rename(Species = species, Model = formula, K = nPars)

#table.clean[table.clean$Model == "~1 ~ 1 ~ 1 ~ 1"] = paste(greeks("epsilon"), "(.), ", greeks("rho"), "(.)", sep = "")

```

#### Table output
```{r}
#write.csv(table.clean, file = "mam_timedep_occ_models_all.csv", row.names = F)

shell.exec("mam_timedep_occ_models_all.csv")

```


# Time-dependent Occupancy Figures

## Detection Probability Models

### Create occupancy model for figures (ε(~year-1), ρ(~year-1))
```{r, warning = FALSE, echo=FALSE}
figure.occ = function(data) {
  mam.dat = format.occ.dat(data)
  survey.years = matrix(c('2016','2018','2020'), 428, 3, byrow = T)

  mam.time.umf = unmarkedMultFrame(y = mam.dat[27:47],
                                      siteCovs = mam.dat[1:26],
                                      yearlySiteCovs = list(year = survey.years),
                                      numPrimary = 3)

  occ = colext(~1, ~1, ~year-1, ~year-1, mam.time.umf)
  
  return(occ)
}

```

### Run occupancy models for figure
```{r, results = "hide", message = FALSE, echo=FALSE}
ra.fig.occ = figure.occ(raccoon)  
op.fig.occ = figure.occ(opossum)    
sq.fig.occ = figure.occ(squirrel)  
cm.fig.occ = figure.occ(cotton.mouse)  
wr.fig.occ = figure.occ(woodrat)  
br.fig.occ = figure.occ(black.rat)

```

### Create function for detection probability plots
```{r, message = FALSE, echo=FALSE}
det.prob.plot = function(model) {
  nd = data.frame(year = c('2016','2018','2020'))
  E.det = predict(model, type = 'det', newdata = nd)
  E.det$year = c(2016,2018,2020)
  
  plot = ggplot(E.det, aes(x = year, y = Predicted)) + 
    geom_line(size = 0.5) +
    geom_point(size = 2.0)+ 
    geom_errorbar(aes(ymin = lower, ymax = upper), size = 0.5, width = 0.1) +
    scale_x_continuous(breaks = seq(2016, 2020, 2)) +
    ylim(0,0.6)+
    ylab(NULL)+
    xlab(NULL)+
    theme_classic()+
    theme(
      aspect.ratio = 0.8,
      axis.line = element_line(size = 0.75),
      axis.text.x = element_blank(),
      axis.text.y = element_blank())
  
  return(plot)
}  

```

```{r}
det.probs = function(model) {
  nd = data.frame(year = c('2016','2018','2020'))
  E.det = predict(model, type = 'det', newdata = nd)
  E.det$year = c(2016,2018,2020)
  
  return(E.det)
} 

det.table = rbind(det.probs(ra.fig.occ), det.probs(op.fig.occ), det.probs(sq.fig.occ), det.probs(cm.fig.occ), det.probs(wr.fig.occ), det.probs(br.fig.occ))

mam.species = c(rep("raccoon", 3), rep("opossum", 3), 
             rep("squirrel", 3), rep("cotton.mouse", 3),  
             rep("woodrat", 3), rep("black.rat", 3))
det.table$species = mam.species
det.table

write.csv(det.table, file = "detection.prediction.table.csv", row.names = F)
shell.exec("detection.prediction.table.csv")

```
### Create detection probability plots
```{r, echo=FALSE}
ra.det.plot = det.prob.plot(ra.fig.occ) +
                theme(
                  axis.text.y = element_text(size = 11, color = "black")
                )

op.det.plot = det.prob.plot(op.fig.occ) +
                theme(
                  axis.text.y = element_text(size = 11, color = "black")
                )

sq.det.plot = det.prob.plot(sq.fig.occ) +
                theme(
                  axis.text.y = element_text(size = 11, color = "black"),
                  axis.text.x = element_text(size = 11, color = "black")
                )
cm.det.plot = det.prob.plot(cm.fig.occ)

wr.det.plot = det.prob.plot(wr.fig.occ)

br.det.plot = det.prob.plot(br.fig.occ) +
                theme(
                  axis.text.x = element_text(size = 11, color = "black")
                )


```


### Combine graphs into one figure
```{r, warning = FALSE, echo=FALSE}
det.fig = ggarrange(ra.det.plot, cm.det.plot, op.det.plot, wr.det.plot, sq.det.plot, br.det.plot, 
            labels = c("RA", "CM", "OP", "WR", "SQ", "BR"), hjust = 0.2, 
            ncol = 2, nrow = 3, align = "hv", font.label = list(color = "black", face = "bold", size = 9))
det.fig = annotate_figure(det.fig, left = text_grob("Detection Probability\n", 
            size = 14, face = "bold", rot = 90), bottom = text_grob("\nYear", size = 14, face = "bold"))
det.fig

```


## Extinction Probability Plots

### Create function for extinction probability plots
```{r, message=FALSE, echo=FALSE}
ext.prob.plot = function(model) {
  nd = data.frame(year = c('2016','2018'))
  E.ext = predict(model, type = 'ext', newdata = nd)
  E.ext$year = c(2016,2018)
  
  plot = ggplot(E.ext, aes(x = year, y = Predicted)) + 
    geom_line(size = 0.5) +
    geom_point(size = 2)+ 
    geom_errorbar(aes(ymin = lower, ymax = upper), size = 0.5, width = 0.1) +
    scale_x_continuous(breaks = seq(2016, 2018, by = 2)) +
    scale_y_continuous(breaks = seq(0, 1, by = 0.5)) +
    ylab(NULL)+
    xlab(NULL)+
    #ylim(0, 1)+
    theme_classic()+
    theme(
      aspect.ratio = 0.8,
      axis.line = element_line(size = 0.75),
      axis.text.x = element_blank()
      #axis.text.y = element_blank()
    )
  
  return(plot)
} 

```

### Create extinction probability plots
```{r, message = FALSE, echo=FALSE}
ra.ext.plot = ext.prob.plot(ra.fig.occ) +
                scale_y_continuous(breaks = seq(0, 1, 0.5)) +
                theme(
                  axis.text.y = element_text(size = 11, color = "black")
                )

op.ext.plot = ext.prob.plot(op.fig.occ) +
                theme(
                  axis.text.y = element_text(size = 11, color = "black")
                )

sq.ext.plot = ext.prob.plot(sq.fig.occ) +
                theme(
                  axis.text.y = element_text(size = 11, color = "black"),
                  axis.text.x = element_text(size = 11, color = "black")
                )

cm.ext.plot = ext.prob.plot(cm.fig.occ)


wr.ext.plot = ext.prob.plot(wr.fig.occ)


br.ext.plot = ext.prob.plot(br.fig.occ) +
                theme(
                  axis.text.x = element_text(size = 11, color = "black")
                )

```

### Combine graphs into one figure
```{r, warning = FALSE, echo=FALSE}
ext.fig = ggarrange(ra.ext.plot, cm.ext.plot, op.ext.plot, wr.ext.plot, sq.ext.plot, br.ext.plot, 
            labels = c("RA", "CM", "OP", "WR", "SQ", "BR"), hjust = 0.2, 
            ncol = 2, nrow = 3, align = "hv", font.label = list(color = "black", size = 9))
ext.fig = annotate_figure(ext.fig, left = text_grob("Extinction Probability\n", 
            size = 14, face = "bold", rot = 90), bottom = text_grob("\nYear", size = 14, face = "bold"))
ext.fig

```

```{r}
ext.probs = function(model) {
  nd = data.frame(year = c('2016','2018'))
  E.ext = predict(model, type = 'ext', newdata = nd)
  E.ext$year = c(2016,2018)
  
  return(E.ext)
} 

ext.table = rbind(ext.probs(ra.fig.occ), ext.probs(op.fig.occ), ext.probs(sq.fig.occ), ext.probs(cm.fig.occ), ext.probs(wr.fig.occ), ext.probs(br.fig.occ))

mam.species = c(rep("raccoon", 2), rep("opossum", 2), 
             rep("squirrel", 2), rep("cotton.mouse", 2),  
             rep("woodrat", 2), rep("black.rat", 2))
ext.table$species = mam.species
ext.table


write.csv(ext.table, file = "extinction.prediction.table.csv", row.names = F)
shell.exec("extinction.prediction.table.csv")

```

